Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-05-20 jss/EpiCompare-paper/EpiCompare-paper.Rmd
section 3 done except for final subsection

In this section, we present the tools implemented in \pkg{EpiCompare} and explain how they aid in the  data analysis pipeline.  In Fig. \ref{fig:pipeline2}, we show how our package's functions fit into the data analysis pipeline introduced in Fig. \ref{fig:pipeline}.  All front-facing functions in \pkg{EpiCompare} are aimed to be as user-friendly as possible. We also focus on providing the user "tidyverse" style functions, that encourage piping objects from one function to the next and follow clear "verb" naming schemes \citep{Wickham2019}. Although users can incorporate \pkg{EpiCompare} into any step in the data analysis pipeline, there are two primary points of entry.  The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation. Figure \ref{fig:pipeline2} captures these different paths, and we highlight how to leverage \pkg{EpiCompare} functionalities in the subsections below.

**Data Pre-processing**

The first step of most data analysis is "cleaning" the raw data so it can be explored. Before data can be explored, they must be collected. Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information. Other data collections focus on aggregate counts of individuals in each epidemic state.  In fact, many times only the number of new infections at each time step (e.g. weekly case counts) is observed.  Compartment totals (amounts of individuals in each state) are then imputed from those case counts other information about the disease and the population of interest.  In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population lens, we provide a function called `agents_to_aggregate()`. This function transforms data about individual/agents' initial entry into each state (e.g. start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time. There are often situations where grouping agents into subpopulations (e.g. subpopulations defined by age or sex) can highlight different aggregate level trends. For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations. In \pkg{EpiCompare}, we facilitate subpopulation analysis\footnote{\shannon{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}} by extending `agents_to_aggregate()` to work with \pkg{dplyr}::`group_by()`, the tidyverse approach to the split-apply-combine paradigm for data analysis \citep{wickham2011}.

[[\textcolor{red}{OLD SENTENCE}
by combining the function \pkg{dplyr}::`group_by()` and `agents_to_aggregate()` to provide aggregation at a subpopulation level\footnote{\ben{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\shannon{agreed. aggregation at a group level}}.
]]

@@

**Exploratory Data Analysis (EDA)**

 In the early stages of a project, familiarizing oneself with the data usually means figuring out useful combinations of visualizations and numerical summaries of the data both at population and subpopulation level.  An expert coder can start with `agents_to_aggregate()` to successfully accomplish exploratory data analysis (EDA) in many ways. \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there are three unique epidemiological states (like in the SIR model). Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}. The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.  Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).

**Model Fitting and Simulations**

After getting a sense of what a past or current epidemic looks like with EDA, the next step in the data analysis pipeline is often model fitting and/or simulation. In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}. While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models. These functions simulate individual-level information based on practitioner estimated transition rates between states and can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.  The function `simulate_SIR_agents()` simulates a basic SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.  This function allows for easy access to SIR model analysis and comparison.  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.  This flexibility in states can be used to also reflect group-based dynamics.  In turn, this allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail. For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.

@@

\shannon{[[NEW TEXT}
If practitioners wish to compare models-to-observations or even models-to-models, they need to post-process their models and simulations to disseminate the results in an easily digestible format. In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables. In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{\shannon{new signpost.  There are two paragraphs below corresponding to the two points.  The second paragraph could use another sentence maybe.  I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}

@@

Because the fortified output contains temporal information and we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations. These filaments can then be fairly compared to one another or passed to further time-invariant analysis tools described below.
\shannon{]]}

[[\textcolor{red}{OLD TEXT BELOW}

\footnote{\ben{[Ben says: I found the structure of this section slightly confusing. It appears that the 2 subsections are (1) model based transformation and (2) simulation based transformations. The fact that there are 2 points isn't quickly apparent when reading it, and I found it odd that the 3rd sentence of the first paragraph introduces model processing (which the second paragraph also discusses).]}\shannon{agreed that this section needs some work.  but to me those aren't the clear points, it seems like (1) put external model results in the same format and (2) transform model output to filaments.  that said i don't see how we're over-selling the package.  would 'transform and disseminate' be better?}}
If practitioners wish to compare models-to-observations or even models-to-models, they needs to post-process their models and simulations to disseminate the results in an easily digestible format. In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.

Model output summaries can be very complex, and as a result, a number of epidemic modeling \proglang{R} packages return a special class.  The special classes often contain a plethora of information about residuals, model diagnostics, input parameters, and more.  While incredibly useful, these special classes can be difficult for novice coders to handle. \shannon{motivates for fortify but not filamental compression}

To this end, \pkg{EpiCompare} provides a series of `fortify`-style methods, called `fortify_aggregate()` which transform output from infectious disease modeling and simulation packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.  These fortify functions have output that is consistent with that of `agents_to_aggregate()`.

To utilize simulations of epidemics in  time-invariant analysis, we also provide a function to convert temporally defined epidemics to filamental representations. Specifically, we provide the function `filament_compression()` to convert simulation(s) to filamental trajectories as expressed by presenting the epidemic as an ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space\footnote{\ben{[Ben says: there's a lot going on in this sentence, and the phrase "as expressed by presenting the epidemic as an ordering of some common fixed number of points" is between the subject and our verb "so that".]}}. These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis are already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.\footnote{\ben{[Ben says: This sentence still doesn't seem to fit into the paragraph or section too well. First, given it's in this paragraph: is it just the simulation based "tools"? I might suggest moving it ot the first introductory sentence - but probably needs some word-smithing as well - I feel like it might be overselling the package a bit.]}}

]]

**Comparisons and Assessment**
\shannon{[[ NEW TEXT}

Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare different models, and simulations to one another. In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g. mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{\shannon{a bit of an awk transition but I wanted to use past wording about TIA.}}  We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of an epidemic has occurred.

>>>>

In this section, we present the tools implemented in \pkg{EpiCompare} and explain how they aid in the  data analysis pipeline.  In Figure \ref{fig:pipeline2}, we show how our package's functions fit into the data analysis pipeline introduced in Figure \ref{fig:pipeline}.  All front-facing functions in \pkg{EpiCompare} are aimed to be as user-friendly as possible. We also focus on providing the user "tidyverse" style functions, that encourage piping objects from one function to the next and follow clear "verb" naming schemes \citep{Wickham2019}. Although users can incorporate \pkg{EpiCompare} into any step in the data analysis pipeline, there are two primary points of entry.  The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation. Figure \ref{fig:pipeline2} captures these different paths, and we highlight how to leverage \pkg{EpiCompare} functionalities in the subsections below.

**Data pre-processing**

The first step of most data analysis is "cleaning" the raw data so it can be explored. Before data can be explored, they must be collected. Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information. Other data collections focus on aggregate counts of individuals in each epidemic state.  In fact, many times only the number of new infections at each time step (e.g. weekly case counts) is observed.  In this setting, compartment totals (amounts of individuals in each state) are then imputed from those case counts and using other information about the disease and the population of interest.  In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population lens, we provide a function called `agents_to_aggregate()`. This function transforms data about individual/agents' initial entry into each state (e.g. start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time. Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregation, conditional on subpopulations (e.g. subpopulations defined by age or sex). By combining the function \pkg{dplyr}::`group_by()` and `agents_to_aggregate()`, \pkg{EpiCompare} provides group level aggregation.

@@

**Exploratory data analysis (EDA)**

 In the early stages of a project, familiarizing oneself with the data usually means figuring out useful combinations of visualizations and numerical summaries of the data both at population and subpopulation level.  An expert coder can start with `agents_to_aggregate()` to successfully accomplish exploratory data analysis (EDA) in many ways. \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there are three unique epidemiological states (like in the SIR model). Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}. The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.  Visualization tools for three-state models were developed because SIR models are some of the most common and basic epidemic state-based models and our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).

**Model fitting and simulations**

After getting a sense of what a past or current epidemic looks like with EDA, the next step in the data analysis pipeline is often model fitting and/or simulation.  While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models. These functions simulate individual-level information based on practitioner estimated transition rates between states and can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.  The function `simulate_SIR_agents()` simulates a basic SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.    Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.  This flexibility in states can be used to also reflect group-based dynamics.  Both of these functions allow users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail. For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.

@@

If practitioners wish to compare models-to-observations or even models-to-models, they need to post-process their models and simulations to disseminate the results in an easily digestible format.  In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.

@@

Because epidemic data is stored in a temporal way, we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations. These filaments can then be fairly compared to one another or passed to further time-invariant analysis tools described below.

**Comparisons and assessment**
\shannon{[[ NEW TEXT}

\shannon{paragraph re-write thoughts: I changed this paragraph to align with the other sub-sections.  We first talk about the step in the pipeline and then how epicompare fits into that pipeline.  The old paragraph skipped the general plots, tables, and 'standard' tools and went straight into filaments.  I re-wrote the paragraph to more clearly indicate that we don't think filaments are the only tools for comparison and assessment.}

Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare \shannon{and assess} different models, and simulations to one another\footnote{\ben{[Ben says: I'm confused why this is just about comparison and not assessment. Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}. In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g. mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{\shannon{a bit of an awk transition but I wanted to use past wording about TIA.}}  We have found these tools to be specifically applicable for situations where only one season or cycle of an epidemic has occurred.

### 2021-05-20 jss/EpiCompare-paper/EpiCompare-paper.tex
section 3 done except for final subsection

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Fig.
\ref{fig:pipeline2}, we show how our package's functions fit into the
data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
front-facing functions in \pkg{EpiCompare} are aimed to be as
user-friendly as possible. We also focus on providing the user
``tidyverse'' style functions, that encourage piping objects from one
function to the next and follow clear ``verb'' naming schemes
\citep{Wickham2019}. Although users can incorporate \pkg{EpiCompare}
into any step in the data analysis pipeline, there are two primary
points of entry. The first point of entry is the very beginning with
pre-processing and visualizing raw data, and the second point of entry
is after modeling and simulation. Figure \ref{fig:pipeline2} captures
these different paths, and we highlight how to leverage \pkg{EpiCompare}
functionalities in the subsections below.

\textbf{Data Pre-processing}

The first step of most data analysis is ``cleaning'' the raw data so it
can be explored. Before data can be explored, they must be collected.
Sometimes individual records are collected, with times of different
states of the epidemic (infection, recovery, etc.) as well as individual
information like network structure, location, and sub-population
information. Other data collections focus on aggregate counts of
individuals in each epidemic state. In fact, many times only the number
of new infections at each time step (e.g.~weekly case counts) is
observed. Compartment totals (amounts of individuals in each state) are
then imputed from those case counts other information about the disease
and the population of interest. In \pkg{EpiCompare}, we focus on
understanding the overall impact of an outbreak at the
aggregate/population level, which allows for streamlined examination of
overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population
lens, we provide a function called \texttt{agents\_to\_aggregate()}.
This function transforms data about individual/agents' initial entry
into each state (e.g.~start of infection, start of recovery, etc.) to an
aggregate view of how many individuals were in a state at a given time.
There are often situations where grouping agents into subpopulations
(e.g.~subpopulations defined by age or sex) can highlight different
aggregate level trends. For example, research by
\citet{rvachev1985,anderson1992,worby2015} develop state-based models
that account for differing disease dynamics in different subpopulations.
In \pkg{EpiCompare}, we facilitate subpopulation
analysis\footnote{\textcolor{orange}{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}}
by extending \texttt{agents\_to\_aggregate()} to work with
\pkg{dplyr}::\texttt{group\_by()}, the tidyverse approach to the
split-apply-combine paradigm for data analysis \citep{wickham2011}.

{[}{[}\textcolor{red}{OLD SENTENCE} by combining the function
\pkg{dplyr}::\texttt{group\_by()} and \texttt{agents\_to\_aggregate()}
to provide aggregation at a subpopulation
level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\textcolor{orange}{agreed. aggregation at a group level}}.{]}{]}

@@

\textbf{Exploratory Data Analysis (EDA)}

In the early stages of a project, familiarizing oneself with the data
usually means figuring out useful combinations of visualizations and
numerical summaries of the data both at population and subpopulation
level. An expert coder can start with \texttt{agents\_to\_aggregate()}
to successfully accomplish exploratory data analysis (EDA) in many ways.
\pkg{EpiCompare} also includes tools that allow a novice coder to
rapidly explore data, provided there are three unique epidemiological
states (like in the SIR model). Building on \pkg{ggplot2} and
\pkg{ggtern} packages, \pkg{EpiCompare}'s \texttt{geom\_aggregate()}
provides a way to explore how different subpopulations experience of an
epidemic \citep{Wickham2016, Hamilton2018}. The function
\texttt{geom\_aggregate()} provides a visualization tool to holistically
examine aggregate level information across different subpopulations by
visualizing each subpopulation's epidemic trajectory in the
three-dimensional state space. Visualization tools for three-state
models were developed because (1) SIR models are some of the most common
and basic epidemic state-based models and (2) our three-dimensional
simplex representation of these epidemics emphasizes a time-invariant
representation of the data (for a refresher see Section
\ref{sec:time-invariant}).

\textbf{Model Fitting and Simulations}

After getting a sense of what a past or current epidemic looks like with
EDA, the next step in the data analysis pipeline is often model fitting
and/or simulation. In this step and the next step (post-processing), we
highlight how practitioners can pair models and simulations of epidemics
from outside of \pkg{EpiCompare} with analysis and simulation tools in
\pkg{EpiCompare}. While \pkg{EpiCompare} does not focus on fitting
models to data, we do provide some flexible functions for simulation of
basic discrete-time epidemic-state models. These functions simulate
individual-level information based on practitioner estimated transition
rates between states and can be combined with
\texttt{agents\_to\_aggregate()} to view these simulations through an
aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
a basic SIR epidemic with user inputs for the number of simulations, the
initial number in each state, the infection and recovery parameters
\((\beta, \gamma\)), and the total number of discrete time steps. This
function allows for easy access to SIR model analysis and comparison.
Beyond SIR models, the function \texttt{simulate\_agents()} takes as
input a user-specified state-transition matrix and other epidemic
parameters to allow the user to create simulations for an outbreak with
\textit{any} number of states and any number of transitions among them.
This flexibility in states can be used to also reflect group-based
dynamics. In turn, this allows for users to explore the space of models
in an intuitive way without getting bogged down by too much mathematical
detail. For consistency, we have made output from
\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
compatible with \texttt{agents\_to\_aggregate()} so aggregate
information may easily be accessed.

@@

If practitioners wish to compare models-to-observations or even
models-to-models, they need to post-process their models and simulations
to disseminate the results in an easily digestible format. In general,
post-processing of modeling and simulation consists of making summary
statistics, plots, and tables. In \pkg{EpiCompare}, we provide (1)
functions to standardize simulation and model output from external
packages and (2) a function to transform standardized simulation and
model output into a format amenable to time-invariant
analysis.\footnote{\textcolor{orange}{new signpost.  There are two paragraphs below corresponding to the two points.  The second paragraph could use another sentence maybe.  I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}

@@

Because the fortified output contains temporal information and we
provide the function, \code{filament_compression()}, to transform
temporally defined epidemics to their filamental representations. These
filaments can then be fairly compared to one another or passed to
further time-invariant analysis tools described below.
\textcolor{orange}{]]}

{[}{[}\textcolor{red}{OLD TEXT BELOW}

\footnote{\textcolor{violet}{[Ben says: I found the structure of this section slightly confusing. It appears that the 2 subsections are (1) model based transformation and (2) simulation based transformations. The fact that there are 2 points isn't quickly apparent when reading it, and I found it odd that the 3rd sentence of the first paragraph introduces model processing (which the second paragraph also discusses).]}\textcolor{orange}{agreed that this section needs some work.  but to me those aren't the clear points, it seems like (1) put external model results in the same format and (2) transform model output to filaments.  that said i don't see how we're over-selling the package.  would 'transform and disseminate' be better?}}
If practitioners wish to compare models-to-observations or even
models-to-models, they needs to post-process their models and
simulations to disseminate the results in an easily digestible format.
In general, post-processing of modeling and simulation consists of
making summary statistics, plots, and tables.

Model output summaries can be very complex, and as a result, a number of
epidemic modeling \proglang{R} packages return a special class. The
special classes often contain a plethora of information about residuals,
model diagnostics, input parameters, and more. While incredibly useful,
these special classes can be difficult for novice coders to handle.
\textcolor{orange}{motivates for fortify but not filamental compression}

To this end, \pkg{EpiCompare} provides a series of
\texttt{fortify}-style methods, called \texttt{fortify\_aggregate()}
which transform output from infectious disease modeling and simulation
packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
which contain information about the total number of individuals in each
state at a given time, for a given simulation. These fortify functions
have output that is consistent with that of
\texttt{agents\_to\_aggregate()}.

To utilize simulations of epidemics in time-invariant analysis, we also
provide a function to convert temporally defined epidemics to filamental
representations. Specifically, we provide the function
\texttt{filament\_compression()} to convert simulation(s) to filamental
trajectories as expressed by presenting the epidemic as an ordering of
some common fixed number of points so that they are equally spaced along
the original path in the proportional state
space\footnote{\textcolor{violet}{[Ben says: there's a lot going on in this sentence, and the phrase "as expressed by presenting the epidemic as an ordering of some common fixed number of points" is between the subject and our verb "so that".]}}.
These tools were developed to provide another natural entry point into
the \pkg{EpiCompare} data analysis pipeline for situations where
modeling and analysis are already completed and practitioners are
looking for simple and transparent tools to help understand and
disseminate
results.\footnote{\textcolor{violet}{[Ben says: This sentence still doesn't seem to fit into the paragraph or section too well. First, given it's in this paragraph: is it just the simulation based "tools"? I might suggest moving it ot the first introductory sentence - but probably needs some word-smithing as well - I feel like it might be overselling the package a bit.]}}

{]}{]}

\textbf{Comparisons and Assessment} \textcolor{orange}{[[ NEW TEXT}

Finally, the data analysis pipeline often ends with plots, tables, and
summary statistics that are used to compare different models, and
simulations to one another. In \pkg{EpiCompare} we provide a set of
comparison and assessment tools for model and simulation results that
extend beyond the standard performance metrics (e.g.~mean squared error
or AIC) and into the lens of time-invariant
analysis.\footnote{\textcolor{orange}{a bit of an awk transition but I wanted to use past wording about TIA.}}
We have found these tools to be specifically applicable for situations
where only one ``season'' or ``cycle'' of an epidemic has occurred.

>>>>

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Figure
\ref{fig:pipeline2}, we show how our package's functions fit into the
data analysis pipeline introduced in Figure \ref{fig:pipeline}. All
front-facing functions in \pkg{EpiCompare} are aimed to be as
user-friendly as possible. We also focus on providing the user
``tidyverse'' style functions, that encourage piping objects from one
function to the next and follow clear ``verb'' naming schemes
\citep{Wickham2019}. Although users can incorporate \pkg{EpiCompare}
into any step in the data analysis pipeline, there are two primary
points of entry. The first point of entry is the very beginning with
pre-processing and visualizing raw data, and the second point of entry
is after modeling and simulation. Figure \ref{fig:pipeline2} captures
these different paths, and we highlight how to leverage \pkg{EpiCompare}
functionalities in the subsections below.

\textbf{Data pre-processing}

The first step of most data analysis is ``cleaning'' the raw data so it
can be explored. Before data can be explored, they must be collected.
Sometimes individual records are collected, with times of different
states of the epidemic (infection, recovery, etc.) as well as individual
information like network structure, location, and sub-population
information. Other data collections focus on aggregate counts of
individuals in each epidemic state. In fact, many times only the number
of new infections at each time step (e.g.~weekly case counts) is
observed. In this setting, compartment totals (amounts of individuals in
each state) are then imputed from those case counts and using other
information about the disease and the population of interest. In
\pkg{EpiCompare}, we focus on understanding the overall impact of an
outbreak at the aggregate/population level, which allows for streamlined
examination of overall trends of an epidemic.

To help the practitioner examine epidemics from an aggregate/population
lens, we provide a function called \texttt{agents\_to\_aggregate()}.
This function transforms data about individual/agents' initial entry
into each state (e.g.~start of infection, start of recovery, etc.) to an
aggregate view of how many individuals were in a state at a given time.
Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often
are interesting in more granular trends that can be detected by
aggregation, conditional on subpopulations (e.g.~subpopulations defined
by age or sex). By combining the function
\pkg{dplyr}::\texttt{group\_by()} and \texttt{agents\_to\_aggregate()},
\pkg{EpiCompare} provides group level aggregation.

@@

\textbf{Exploratory data analysis (EDA)}

In the early stages of a project, familiarizing oneself with the data
usually means figuring out useful combinations of visualizations and
numerical summaries of the data both at population and subpopulation
level. An expert coder can start with \texttt{agents\_to\_aggregate()}
to successfully accomplish exploratory data analysis (EDA) in many ways.
\pkg{EpiCompare} also includes tools that allow a novice coder to
rapidly explore data, provided there are three unique epidemiological
states (like in the SIR model). Building on \pkg{ggplot2} and
\pkg{ggtern} packages, \pkg{EpiCompare}'s \texttt{geom\_aggregate()}
provides a way to explore how different subpopulations experience of an
epidemic \citep{Wickham2016, Hamilton2018}. The function
\texttt{geom\_aggregate()} provides a visualization tool to holistically
examine aggregate level information across different subpopulations by
visualizing each subpopulation's epidemic trajectory in the
three-dimensional state space. Visualization tools for three-state
models were developed because SIR models are some of the most common and
basic epidemic state-based models and our three-dimensional simplex
representation of these epidemics emphasizes a time-invariant
representation of the data (for a refresher see Section
\ref{sec:time-invariant}).

\textbf{Model fitting and simulations}

After getting a sense of what a past or current epidemic looks like with
EDA, the next step in the data analysis pipeline is often model fitting
and/or simulation. While \pkg{EpiCompare} does not focus on fitting
models to data, we do provide some flexible functions for simulation of
basic discrete-time epidemic-state models. These functions simulate
individual-level information based on practitioner estimated transition
rates between states and can be combined with
\texttt{agents\_to\_aggregate()} to view these simulations through an
aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
a basic SIR epidemic with user inputs for the number of simulations, the
initial number in each state, the infection and recovery parameters
\((\beta, \gamma\)), and the total number of discrete time steps. Beyond
SIR models, the function \texttt{simulate\_agents()} takes as input a
user-specified state-transition matrix and other epidemic parameters to
allow the user to create simulations for an outbreak with \textit{any}
number of states and any number of transitions among them. This
flexibility in states can be used to also reflect group-based dynamics.
Both of these functions allow users to explore the space of models in an
intuitive way without getting bogged down by too much mathematical
detail. For consistency, we have made output from
\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
compatible with \texttt{agents\_to\_aggregate()} so aggregate
information may easily be accessed.

@@

If practitioners wish to compare models-to-observations or even
models-to-models, they need to post-process their models and simulations
to disseminate the results in an easily digestible format. In
\pkg{EpiCompare}, we provide (1) functions to standardize simulation and
model output from external packages and (2) a function to transform
standardized simulation and model output into a format amenable to
time-invariant analysis.

@@

Because epidemic data is stored in a temporal way, we provide the
function, \code{filament_compression()}, to transform temporally defined
epidemics to their filamental representations. These filaments can then
be fairly compared to one another or passed to further time-invariant
analysis tools described below.

\textbf{Comparisons and assessment} \textcolor{orange}{[[ NEW TEXT}

\textcolor{orange}{paragraph re-write thoughts: I changed this paragraph to align with the other sub-sections.  We first talk about the step in the pipeline and then how epicompare fits into that pipeline.  The old paragraph skipped the general plots, tables, and 'standard' tools and went straight into filaments.  I re-wrote the paragraph to more clearly indicate that we don't think filaments are the only tools for comparison and assessment.}

Finally, the data analysis pipeline often ends with plots, tables, and
summary statistics that are used to compare
\textcolor{orange}{and assess} different models, and simulations to one
another\footnote{\textcolor{violet}{[Ben says: I'm confused why this is just about comparison and not assessment. Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}.
In \pkg{EpiCompare} we provide a set of comparison and assessment tools
for model and simulation results that extend beyond the standard
performance metrics (e.g.~mean squared error or AIC) and into the lens
of time-invariant
analysis.\footnote{\textcolor{orange}{a bit of an awk transition but I wanted to use past wording about TIA.}}
We have found these tools to be specifically applicable for situations
where only one season or cycle of an epidemic has occurred.
